### 2017-07-03 reasoning.tex
Sec. 4 part 1.

% We now describe a proof system that lets us demonstrate that a \txnimp
% program satisfies its high-level invariants in all small-step
% executions that satisfy a chosen trace invariant ($\I$).\footnote{It is
% important to keep in mind the distinction between high-level invariants
% (\eg~\C{X$\ge$0}) and trace invariants (\eg~$\I =
% \lambda\E.\,\C{RC}(\C{Wd1})\wedge \C{SI}(\C{Wd2})$). The former
% constitute \emph{proof} \emph{obligations} for the programmer, whereas
% the latter is a combination of transaction-specific \emph{assumptions} about 
% operational characteristics of the underlying data store. } 
% Since the trace invariant captures transaction-isolation constraints,
% the demonstration is a proof that the given selection of isolation
% levels for transactions is sufficient to enforce a program's invariants.

% As mentioned earlier, our proof system is a rely-guarantee framework
% that admits compositional reasoning by abstracting away environmental
% interference as a rely relation. A conspicuous difference between a
% standard development of rely-guarantee and ours is that, while the
% former reasons in terms of program states (variable to value
% bindings), we reason in terms of executions as captured by their
% traces ($\E$). In particular, our rely ($R$) and guarantee ($G$)
% relations relate executions (i.e., $R,\,G\,\subseteq\,\E\times\E$),
% and our pre- ($P$) and post- ($Q$) conditions are assertions over
% executions (i.e., $P,\,Q\, : \E \rightarrow \Prop$). Our development
% also facilitates state-based reasoning via the interpretation function
% ($\interp{\cdot}$) introduced in \S\ref{sec:syntax}, which interprets
% the given set of effects as a state. For example, if a \emph{bi-state}
% rely relation relates every pair of states $\sigma$ and $\sigma'$ such
% that $\sigma'(X) \ge \sigma(X) \ge 0$, the corresponding
% \emph{bi-execution} rely relation relates every pair of executions
% $\E$ and $\E'$ such that $\interp{\E.\A}(X) \ge \interp{\E'.\A}(X) \ge
% 0$ . Assertions on states are also written similarly. For instance,
% consider the post-condition $\C{B\,=\,k-a1-a2}$ of the program in
% Fig.~\ref{fig:motiv-eg-1}. The corresponding assertion on the
% post-state ($\lambda\sigma.~\sigma(\C{B}) \,=\, \C{k-a1-a2}$) asserts
% that in all states resulting from executing the program, the value of
% \C{B} is \C{k-a1-a2}. The equivalent execution-based assertion
% ($\lambda\E.~\interp{\E.\A}(\C{B})\,=\,\C{k-a1-a2}$) asserts that in
% all executions of the program, the value of \C{B} is \C{k-a1-a2}.
% However, having access to an execution facilitates assertions that go
% beyond the state. An example is an invariant of \C{Wd1} described
% abstractly in \S\ref{sec:motivation}, and reified as an
% execution-based assertion below:

% \begin{smathpar}
% \begin{array}{l}
%   \lambda\E.~ \neg(\underE{\committed(\C{Wd1}}) \conj 
%       \underE{\C{Wd1} \wrstoar \C{B}}\\
%   \hspace*{0.2in}\conj(\neg(\underE{\committed(\C{Wd2})}) \Rightarrow 
%       \C{\interp{\E.\A}(B) = k-a1}) \\
%    \hspace*{0.2in}\conj (\underE{\committed(\C{Wd2})} \Rightarrow 
%       \C{\interp{\E.\A}(B) = k-a1-a2})
% \end{array}
% \end{smathpar}
% % Where, $\committed$ is defined as:
% % \begin{smathpar}
% % \begin{array}{lcl}
% % \underE{\committed(T_i)} & \Leftrightarrow &
% %   \exists\eta.~\underE{\eta\in T_i} \conj \kind(\eta) = \C{COMMIT}\\
% % \end{array}
% % \end{smathpar}

% \noindent One of the conjuncts asserts that \C{Wd1} has written to \C{B} - a
% fact which cannot be deduced solely from the value of \C{B} (esp. in the
% presence of interference), but can be expressed as a proposition over
% the execution trace.

% \subsection{The Rely-Guarantee Judgment}
% \label{sec:rely-guarantee}

@@

% The standard rely-guarantee judgment is a quintuple
% $\rg{P,R}{c}{G,Q}$, which informally asserts that if a command $c$ is
% executed in a state that satisfies its pre-condition $P$, provided
% that every interference step during the execution is contained inside
% the rely relation $R$, then the effect that each step of executing $c$
% has on the state is captured by $G$, and the final state of execution
% satisfies the post-condition $Q$. The terms ``execution'' and ``step''
% are defined with respect to an operational semantics, which, in our
% case, is parameterized on the trace invariant $\I$. Consequently, our
% rely-guarantee judgment for commands is the sextuple
% $\rg{\I,P,R}{c}{G,Q}$, whose semantics differs from the quintuple in
% that it requires every step of the execution to additionally preserve
% the trace invariant $\I$ for the post-condition to be valid.  As
% usual, if $c$ is inside a transaction $T_i$, we write $\txnbox{c}_i$.
% \txnimp expressions are side-effecting (they generate \C{RD} effects),
% and admit interference during their evaluation.  Therefore, like
% commands, expressions need to be reasoned about explicitly. But,
% unlike commands, expressions evaluate to values, and the reasoning
% framework should also admit assertions on such values.  We therefore
% define a separate judgment for expressions - the septuple
% $\rg{\I,P,R}{e}{G,C,Q}$. The new entrant $C : \mathbb{N} \rightarrow
% \Prop$ is an assertion on the return value of $e$.

>>>>

We now describe a proof system that lets us prove the correctness of a \txnimp
program $c$ w.r.t its high-level invariant $I$, on a machine that
satisfies the isolation specifications ($\I$) of its
transactions\footnote{Note the difference between $I$ and $\I$. The
former constitute \emph{proof} \emph{obligations} for the programmer,
whereas the latter describes a transaction's \emph{assumptions} about
operational characteristics of the underlying machine.}.  Our proof
system is essentially an adapatation of the rely-guarantee reasoning
framework to the setting of weakly isolated database transactions. It
deals with the additional complexity introduced by this setting, while
also taking advantage of the opportunities it provides to simplify the
reasoning. The primary challenge is the weak isolation; how do we
relate a transaction's isolation specification ($\I$) to its rely
relation ($R$) that describes the environment, so that the
interference is considered only insofar as the isolation level allows
it? Our proof system shows how to compute a rely relation ($R$) modulo
the isolation specification ($\I$) of a transaction. Another
characteristic of the transaction setting is the atomicity of a
trasaction's aggregate writes; a transaction's writes are not visible
to its peers until it commits. In the context of rely-guarantee, this
means that the transactions guarantee ($G$) should capture the
aggregate effect of a transaction, and not its individual writes.
While \C{atomic} blocks are also present in the shared-memory
programs, the fact that a transactions are weakly isolated introduces
complexity.  Unlike an \C{atomic} block, the effect of a transaction
is \emph{not} a sequential composition of the effects of its
statements because each statement witnesses a potentially different
version of the state. This is illustrated in
Fig.~\ref{fig:atomic-vs-transaction}, where a series of operations on
\C{x} can be composed to describe the effect of an \C{atomic} block,
while same cannot be done in a transaction. Thus, weak isolation needs
to be taken into account when defining and verifying guarantees, and
our proof system shows how.

\subsection{The Rely-Guarantee Judgment}
\label{sec:rely-guarantee}

@@

Fig.~\ref{fig:rg-rules} shows an illustrative subset of the
rely-guarantee (RG) reasoning rules for $\txnimp$. We define two RG
judgments: top-level ($\rg{I,R}c{G,I}$), and transaction-local ($\R
\vdash \hoare{P}c{Q}$).  Recall that the standard RG judgment is the
quintuple $\rg{P,R}{c}{G,Q}$. Instead of a separate $P$ and $Q$, our
top-level judgment has $I$ for both pre- and post-conditions, because
our focus is on verifying that a \txnimp program \emph{preserves}
database's consistency conditions\footnote{The terms \emph{consistency
condition}, \emph{high-level invaraint}, and \emph{integrity
constraint} all mean the same.}. Transaction-local RG judgment doesn't
include a guarantee relation because transaction-local effects are not
visible outside a transaction. Also, the rely relation ($\R$) of the
transaction-local judgment is not same as the top-level rely relation
($R$) as it takes into account the transaction's isolation
specification ($\I$). Intuitively, $\R$ is $R$ modulo $\I$; formal
definition will be given shortly. Recall that a transaction writes to
its local database ($\stl$), which is then flushed when the
transaction commits. Thus, the guarantee of a transaction depends on
the state of its local database at the commit point. The pre- and
post-condition assertions ($P$ and $Q$) in the local judgment
facilitate tracking the changes to the transaction-local state, which
eventually helps us prove the validity of the transaction's guarantee.
Note that $P$ and $Q$ are bi-state assertions; they relate
transaction-local database state ($\stl$) to the global database state
($\stg$). Thus, the transaction-local judgment effectively tracks how
transaction-local and global states change in relation to each other.

The quintessential aspect of a rely-guarantee judgment is the
stability condition, which, intuitively, requires the validity of an
assertion $\phi$ to be unaffacted by the interference, i.e., the rely
relation $R$. In conventional RG, stability is defined as following
($\sigma$ denotes a state):
\begin{smathpar}
\begin{array}{lcl}
\stable(R,\phi) & \Leftrightarrow & \forall \sigma,\sigma'.~
\phi(\sigma) \conj R(\sigma,\sigma') \Rightarrow \phi(\sigma')\\
\end{array}
\end{smathpar}
Due to the presence of local and global database states, and isolation
specification, we use multiple definitions of stability in
Fig.~\ref{fig:rg-rules}, but they all convey the same intuition as
above. We will introduce them as they are encountered.

\rulelabel{RG-Txn} is the top-level rule that lets us prove a
transaction preserves the high-level invariant $I$ when executed under
the required isolation as specified by $\I$. It relies on the
transaction-local judgment to verify the the transaction body ($c$).
The precondition $P$ of $c$ must follow from the fact that the
transaction-local database ($\stl$) is initially empty, and the global
database satisfies the high-level invariant $I$. The rely relation
($\R_l$) for the local judgment is a ternary relation computed as $R$
modulo $\I$ as following:
\begin{smathpar}
\begin{array}{lcl}
\R_l(\stl,\stg,\stg') & \Leftrightarrow & R(\stg,\stg') \conj
\I\,\C{true}\,(\stl,\stg,\stg')
\end{array}
\end{smathpar}
Thus, $\R_l$ allows an interference only if it doesn't violate the
execution-time isolation specification of the transaction. If a
certain interference violates the isolation spec, i.e., $R(\stg,\stg')
\Rightarrow \neg(\I\,\C{true}\,(\stl,\stg,\stg'))$, then
$\R_l(\stl,\stg,\stg') \Leftrightarrow false$, and any assertion is
trivially stable w.r.t that interference. This is sensible considering
such interference is pre-empted in the operational semantics.

Recall that $P$ and $Q$ of the transaction-local RG judgment are
binary assertions; they relate local and global database states. The
local judgment rules require one or both of them to be stable. The
stability of a binary assertion $Q$ w.r.t a ternary rely relation $\R$
is defined as following:
\begin{smathpar}
\begin{array}{c}
\forall \stl,\stg,\stg'.~ Q(\stl,\stg) \conj \R(\stl,\stg,\stg')
\Rightarrow Q(\stl,\stg')
\end{array}
\end{smathpar}
That is, if $Q$ relates $\stl$ to $\stg$, and an interference allowed
by the isolation specfication (which implicitly considers the local
state $\stl$) takes $\stg$ to $\stg'$, the $Q$ must also relate $\stl$
to $\stg'$.

For the guarantee $G$ of the transaction to be valid, it must follow
from the post-condition $Q$ of the body, provided that $Q$ is stable
w.r.t the commit-time interference captured by $R_c$. $R_c$, like
$R_l$, is computed as rely relation modulo isolation, except that
commit-time isolation ($\I\,\C{false}$) is considered. The validity of
$G$ is captured by the following implication:
\begin{smathpar}
\begin{array}{c}
  \forall \stl,\stg.~ Q(\stl,\stg) \Rightarrow G(\stg, \stl \gg \stg)\spc
\end{array}
\end{smathpar}
In other words, if $Q$ relates the transaction-local database state
($\stl$) to the state of the global database ($\stg$) before the
commit, then $G$ must relate the states of the global database before
and after the commit. The act of commit captured by the flush
operation ($\stl\gg\stg$). Once we establish the validity of $G$ as a
faithful representative of the transaction, we can verify that the
transaction preserves the high-level invariant $I$ by checking the
stability of $I$ w.r.t $G$, i.e., $\forall \stg,\stg'.~I(\stg) \wedge
G(\stg,\stg') \Rightarrow I(\stg')$.

A characteristic of RG reasoning is that stability of an assertion is
always proven w.r.t to $R$, and not $R^{*}$, although interference
may include multiple environment steps, and $R$ only captures a single
step. This is nonetheless sound due to the the inductive reasoning: if
$Q$ is preserved by every step of $R$, then $Q$ is preserved by
$R^{*}$, and vice-versa.  However, such reasoning does
extend naturally to isolation-constrained interference ($\R$)
because $R^{*}$ modulo $\I$ is not same as $\R^{*}$; the former is a
transitive relation constrained by $\I$, whereas the latter is the
transitive closure of a $\I$-constrained relation. We therefore
introduce a side-condition on $\I$ that restores the equality. The
condition requires $\I$ to allow an intereference $R^{*}(\stg,\stg'')$,
for two database states $\stg$ and $\stg''$, only if it also allows
interference for every prefix of $R^{*}(\stg,\stg'')$. In other words,
if $\I$ disallows interference from $\stg$ to $\stg'$, then an
$R$-step from $\stg'$ to $\stg''$ shouldn't make the interference from
$\stg$ to $\stg''$ valid. We call this the stability condition on
$\I$, defined as below:
\begin{smathpar}
\begin{array}{lcl}
  \stable(R,\I) & \Leftrightarrow & \forall \stl,\stg,\stg',\stg''.~
  \neg\I(\stl,\stg,\stg') \conj R(\stg',\stg'') \Rightarrow
  \neg\I(\stl,\stg,\stg'')
\end{array}
\end{smathpar}
It can be easily verified that the above stability condition is
satisfied by the isolation axioms from Sec.~\ref{sec:isolation}. For
instance, $\I_{ss}$, the snapshot axiom, is stable because if is
invalid, then an interference has already modified a record, and no
further interference will restore the original record. This is because
records bear transaction ids, and the transaction that wrote the
original record has long committed.
